A free/open-source Kazakh-Tatar machine translation system

### 2013-06-30 paper/paper.tex
explanation

@@

%Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal] (сен, сендер, сіз, сіздер), whereas Tatar collapse all pronouns except the [-plural, -formal] син into one pronoun (сез).  This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
Another example of a far-reaching morphological difference between Tatar and Kazakh is the presence of a 
four-way distinction in Kazakh's 2\super{nd} person system (both pronouns and agreement suffixes), where Tatar only 
has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], 
whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in Table~\ref{tab:2pprons}.
\begin{table}[htbp]
	\centering
%\begin{subtables}
	%\begin{subtable}[l]
	%\subtable[Kazakh 2nd person pronouns]{
	%\begin{subtable}
	\subfloat[Kazakh 2\super{nd} person pronouns]{
		\begin{tabular}{lll}
			\toprule
			 & {\texttt{[-pl]}} & {\texttt{[+pl]}} \\\midrule
			\hspace{-4pt}{\texttt{[-frm]}}\hspace{-4pt} & сен\hspace{-4pt} & сендер\hspace{-4pt} \\
			\hspace{-4pt}{\texttt{[+frm]}}\hspace{-4pt} & сіз\hspace{-4pt} & сіздер\hspace{-4pt} \\\bottomrule
		\end{tabular}
	%	\caption{Kazakh 2nd person pronouns}
	%\end{subtable}
	%\end{subtable}
	%\begin{subtable}
	%\begin{subtable}
	%}\subtable[Tatar 2nd person pronouns]{
	}
	\subfloat[Tatar 2\super{nd} person pronouns]{
		\begin{tabular}{lll}
			\toprule
			 & {\texttt{[-pl]}} & {\texttt{[+pl]}}\hspace{-4pt} \\\midrule
			\hspace{-4pt}{\texttt{[-frm]}}\hspace{-4pt} & син\hspace{-4pt} & сез \\
			\hspace{-4pt}{\texttt{[+frm]}}\hspace{-4pt} & сез\hspace{-4pt} & сез \\\bottomrule
		\end{tabular}%}
	%	\caption{Kazakh 2nd person pronouns}
	%\end{subtable}
	%\end{subtable}
	}
	\caption{2\super{nd} pers.\ pronoun systems of Kazakh and Tatar}
	\label{tab:2pprons}
\end{table}
%\end{subtables}

@@

	}\\
	\subfloat[Tatar past tense morphology]{
		\begin{tabular}{lll}
			\toprule
			 & [$-$new] & [$+$new] \\\midrule
			{}[$+$reliable] & -DI- & -GAн- икән \\
			{}[$-$reliable] & -GAн- & --- \\\bottomrule
		\end{tabular}
%		\begin{tabular}{lll}
%			\toprule
%			 & past \\\midrule
%			eyewitness & -DI- \\
%			non-eyewitness/hearsay & -GAн- \\
%			newly acquired information & -GAн- икән \\\bottomrule
%		\end{tabular}
	}
	\caption{A comparison of the basic past-tense morphology of Kazakh and Tatar}
	\label{tab:pasttense}
\end{table}
%\vspace{-1ex}
The morphotactics of the cognate Kazakh 
distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are 
different.\footnote{This comparison is made without regard to semantic alignment.}  Specifically, in both languages, the person agreement takes the form of a person copula suffix; however
in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``you apparently went''), whereas in 
Tatar this suffix intervenes between the two pieces of the `compound' tense morpheme (e.g., баргансың икән ``I guess you went'').
% I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
% They say you went", i.e. it is closer to Kazakh. "I guess you went" sounds like you are supposing that, which is not the case.
% Maybe I just don't know the whole semantics of the english "to guess". If you are sure about it, just leave it as it is.

@@

\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1ex}
%\a
\label{ex:like-kaz}
\begingl
\gla Маған сөздік түзген ұнайды. //
\glb маған сөздік түз-GAн ұна-E-дI //
\glc \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\xe
%\a\label{ex:like-tat}
\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1ex}
\label{ex:like-tat}
\begingl
\gla Мин сүзлек төзергә яратам. //
\glb мин сүзлек төз-IргA ярат-E-м //
\glc \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1ex}
\label{ex:kerek}
%\a
\begingl
\gla Мен үйге қайтуым керек. //
\glb мен үй-GA қайт-у-Iм керек //
\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
\glft \eng{I need to go home.} //
\endgl
\xe
\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1ex}
%\a
\begingl
\gla Миңа өйгә кайтырга кирәк. //
\glb мин-GA өй-GA кайт-IргA кирәк //
\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
\glft \eng{I need to go home.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1ex}
\label{ex:try}
%\a
\begingl
\gla Айгүл оны табуға әрекет жасап жүр. //
\glb Айгүл о-NI тап-у-GA әрекет жаса-Iп жүр //
\glc Aygül he-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
\glft \eng{Aygül is trying to find him.} //
\endgl
\xe
\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1ex}
%\a
\begingl
\gla Айгөл аны табарга тырыша. //
\glb Айгөл а-NI тап-AргA тырыш-E //
\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
\glft \eng{Aygöl is trying to find him.} //
\endgl
\xe

\vspace{1ex}
As shown in (\ref{ex:GAlI}), the Tatar infinitive also corresponds to a verbal adverb form in Kazakh.
%\vspace{-1ex}

@@

%\a
%\begingl
%\gla Мин син белән сөйлешергә килдем. //
%\glb мин син белән сөйле-ш-IргA кил-DI-м. //
%\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
%\glft \eng{I came to speak with you.} //
%\endgl
\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1ex}
%\a
\begingl
\gla Мин синең белән сөйлешергә килдем. //
\glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
\glc I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe

@@

The output of each morphological analyser is highly ambiguous, measured at around 2.4 morphological analyses per form for Kazakh.  The goal of the CG rules is to select the correct analysis when there are multiple analyses.  As of r43664, the ambiguity was down to 1.4 analyses per form.

@@

@@

The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}  Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.

@@

The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}

>>>>

An MT system between Tatar and Kazakh is potentially of great use to the language communities.  Direct MT can save time and money over going through e.g.\ Russian (and the system is much easier to develop).  Some accompanying language resources (e.g.,\ morphol.\ analysers, disambiguators) can be repurposed.

@@

%Another example of a far-reading morphological difference between Tatar and Kazakh reflects the presence of a four-way distinction in Kazakh's 2nd person system, where Tatar only has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal] (сен, сендер, сіз, сіздер), whereas Tatar collapse all pronouns except the [-plural, -formal] син into one pronoun (сез).  This does not only affect affect the pronoun system: since all finite verb forms morphologically agree in person and number with their subject and all possessed nouns agree in person and number with their possessor (with pronouns used in both circumstances only for emphasis and clarification, as is typical in pro-drop languages), the Kazakh and Tatar systems of agreement suffixes have the same pattern.
Another example of a far-reaching morphological difference between Tatar and Kazakh is the presence of a 
four-way distinction in Kazakh's 2\super{nd} person system (both pronouns and agreement suffixes), where Tatar only 
has a two-way distinction.  Kazakh has a distinct pronoun for all combinations of [±plural, ±formal], 
whereas Tatar collapses all pronouns except the [-plural, -formal] into one pronoun, as summarised in Table~\ref{tab:2pprons}.
\vspace{-1em}
\begin{table}[htbp]
	\centering
%\begin{subtables}
	%\begin{subtable}[l]
	%\subtable[Kazakh 2nd person pronouns]{
	%\begin{subtable}
	\subfloat[Kazakh 2\super{nd} person pronouns]{
		\begin{tabular}{lll}
			\toprule
			 & {\texttt{[-pl]}} & {\texttt{[+pl]}} \\\midrule
			\hspace{-4pt}{\texttt{[-frm]}}\hspace{-4pt} & сен\hspace{-4pt} & сендер\hspace{-4pt} \\
			\hspace{-4pt}{\texttt{[+frm]}}\hspace{-4pt} & сіз\hspace{-4pt} & сіздер\hspace{-4pt} \\\bottomrule
		\end{tabular}
	%	\caption{Kazakh 2nd person pronouns}
	%\end{subtable}
	%\end{subtable}
	%\begin{subtable}
	%\begin{subtable}
	%}\subtable[Tatar 2nd person pronouns]{
	}
	\subfloat[Tatar 2\super{nd} person pronouns]{
		\begin{tabular}{lll}
			\toprule
			 & {\texttt{[-pl]}} & {\texttt{[+pl]}}\hspace{-4pt} \\\midrule
			\hspace{-4pt}{\texttt{[-frm]}}\hspace{-4pt} & син\hspace{-4pt} & сез \\
			\hspace{-4pt}{\texttt{[+frm]}}\hspace{-4pt} & сез\hspace{-4pt} & сез \\\bottomrule
		\end{tabular}%}
	%	\caption{Kazakh 2nd person pronouns}
	%\end{subtable}
	%\end{subtable}
	}
	\caption{2\super{nd} pers.\ pronoun systems of Kazakh and Tatar}
	\label{tab:2pprons}
\end{table}
%\end{subtables}
\vspace{-1ex}

@@

	}\\
	\subfloat[Tatar past tense morphology]{
		\begin{tabular}{lll}
			\toprule
			 & [$-$new] & [$+$new] \\\midrule
			{}[$+$reliable] & -DI- & -GAн- икән \\
			{}[$-$reliable] & -GAн- & --- \\\bottomrule
		\end{tabular}
%		\begin{tabular}{lll}
%			\toprule
%			 & past \\\midrule
%			eyewitness & -DI- \\
%			non-eyewitness/hearsay & -GAн- \\
%			newly acquired information & -GAн- икән \\\bottomrule
%		\end{tabular}
	}
	\caption{A comparison of the basic past-tense morphology of Kazakh and Tatar}
	\label{tab:pasttense}
\end{table}
\vspace{-1ex}
%\vspace{-1ex}
The morphotactics of the cognate Kazakh 
distant non-eywitness past (\texttt{-GAн екен-}) and Tatar newly-acquired-information past (\texttt{-GAн- икән}) are 
different.\footnote{This comparison is made without regard to semantic alignment.}  Specifically, in both languages, the person agreement takes the form of a person copula suffix; however
in Kazakh this suffix follows the tense morphemes (e.g., барған екенсің ``you apparently went''), whereas in 
Tatar this suffix intervenes between the two pieces of the `compound' tense morpheme (e.g., баргансың икән ``I guess you went'').
% I think that the meaning of the Tatar "баргансың икән" is also "It turns out you went, It appears that you went,
% They say you went", i.e. it is closer to Kazakh. "I guess you went" sounds like you are supposing that, which is not the case.
% Maybe I just don't know the whole semantics of the english "to guess". If you are sure about it, just leave it as it is.

@@

\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1.5ex}
%\a
\label{ex:like-kaz}
\begingl
\gla Маған сөздік түзген ұнайды. //
\glb маған сөздік түз-GAн ұна-E-дI //
\glc \gmk{1p.Sg.Dat} dictionary compile-\gmk{Ger} like-\gmk{Aor}-\gmk{3p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\xe
%\a\label{ex:like-tat}
\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1.5ex}
\label{ex:like-tat}
\begingl
\gla Мин сүзлек төзергә яратам. //
\glb мин сүзлек төз-IргA ярат-E-м //
\glc \gmk{1p.Sg} dictionary compile-\gmk{Inf} like-\gmk{Pres}-\gmk{1p.Sg} //
\glft \eng{I like writing dictionaries.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1.5ex}
\label{ex:kerek}
%\a
\begingl
\gla Мен үйге қайтуым керек. //
\glb мен үй-GA қайт-у-Iм керек //
\glb I home-\gmk{Dat} go-\gmk{Ger}-\gmk{1sg} need //
\glft \eng{I need to go home.} //
\endgl
\xe
\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1.5ex}
%\a
\begingl
\gla Миңа өйгә кайтырга кирәк. //
\glb мин-GA өй-GA кайт-IргA кирәк //
\glb I-\gmk{Dat} home-\gmk{Dat} go-\gmk{Inf} need //
\glft \eng{I need to go home.} //
\endgl
\xe

@@

\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1.5ex}
\label{ex:try}
%\a
\begingl
\gla Айгүл оны табуға әрекет жасап жүр. //
\glb Айгүл о-NI тап-у-GA әрекет жаса-Iп жүр //
\glc Aygül he-\gmk{Acc} find-\gmk{Ger}-\gmk{Dat} effort make-\gmk{Prc} \gmk{Prog} //
\glft \eng{Aygül is trying to find him.} //
\endgl
\xe
\pex[everygla=,everyglb=,everyglc=,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1.5ex}
%\a
\begingl
\gla Айгөл аны табарга тырыша. //
\glb Айгөл а-NI тап-AргA тырыш-E //
\glb Aygöl \gmk{3p.Sg}-\gmk{Acc} find-\gmk{Inf} тырыш-\gmk{Pres} //
\glft \eng{Aygöl is trying to find him.} //
\endgl
\xe

%\vspace{1ex}
%\vspace{-1ex}
\vspace{0.8ex}

As shown in (\ref{ex:GAlI}-\ref{ex:GAlI2}), the Tatar infinitive also corresponds to a verbal adverb form in Kazakh.

\vspace{-1ex}

@@

%\a
%\begingl
%\gla Мин син белән сөйлешергә килдем. //
%\glb мин син белән сөйле-ш-IргA кил-DI-м. //
%\glb I you with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
%\glft \eng{I came to speak with you.} //
%\endgl
\pex[everygla=,everyglb=,everyglc=\small,aboveglbskip=0pt,aboveglftskip=0ex]  %glstyle=multilevel,aboveglaskip=0pt
\vspace{-1ex}
\label{ex:GAlI2}
%\a
\begingl
\gla Мин синең белән сөйлешергә килдем. //
\glb мин син-Iң белән сөйле-ш-IргA кил-DI-м. //
\glc I you-\gmk{Gen} with talk-\gmk{Coop}-\gmk{Inf} come-\gmk{Ifi}-\gmk{1p.Sg} //
\glft \eng{I came to speak with you.} //
\endgl
\xe

@@

The output of each morphological analyser is highly ambiguous, measured at around 2.4 morphological analyses per form for Kazakh.  The goal of the CG rules is to select the correct analysis when there are multiple analyses.  Currently, ambiguity is down to 1.4 analyses per form.
%As of r43664, the ambiguity was down to 1.4 analyses per form.

@@

%FIXME NOW
All evaluation was tested against version 0.1.0, or {\tt {\small rNNNNNN}} in the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/trunk/apertium-kaz-tat}}
%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}

@@

%The versions of the transducers tested were {\tt {\small r43595}} from the Apertium SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/staging/apertium-kaz-tat}}
Corpora were divided into 10 parts each; the coverage numbers given are the averages of the calculated percentages of number of words analysed for each of these parts, and the standard deviation presented is the standard deviation of the coverage on each corpus.

@@

The system is available as free/open-source software under the GNU GPL and the whole system may be downloaded
from SVN.\footnote{\url{https://apertium.svn.sourceforge.net/svnroot/apertium/trunk/apertium-kaz-tat}}
